Emergent dual scaling of riverine biodiversity

### 2021-03-23 manuscript_pnas.Rmd
Update manuscript_pnas.Rmd

**Abstract:** A prevailing paradigm suggests that species richness increases with area in a decelerating way; this ubiquitous power-law scaling, the species-area relationship, has formed the foundation of many conservation strategies. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because the scale-invariant complexity of fractal ecosystem structure may drive ecological dynamics in space. Here, we use theory and analysis of extensive fish community data from two distinct geographic regions (Hokkaido Island in Japan and the midwestern United States) to show that ecosystem size and complexity play equally-important roles in scaling biodiversity in rivers. The recurrent merging of various streams and rivers forms fractal branching systems, where the prevalence of branching (ecosystem complexity) represents a macro-scale control of the ecosystem's habitat heterogeneity. In the meantime, ecosystem size dictates metacommunity size and total habitat diversity, two factors regulating biodiversity in nature. Our theory predicted that, regardless of simulated species' traits, larger and more branched 'complex' networks supported greater species richness due to increased space and environmental heterogeneity. Both relationships were linear on a log-log scale, indicating a general power-law scaling along the two orthogonal dimensions of ecosystem structure, i.e., size and complexity. In support of this theoretical prediction, the robust dual power-law scaling of riverine fish diversity has consistently emerged across the study regions despite hosting different fauna with distinct evolutionary histories. The dual scaling of biodiversity may be a pervasive emergent property of branching networks with important implications for biodiversity conservation.

>>>>

**Abstract:** A prevailing paradigm suggests that species richness increases with area in a decelerating way; this ubiquitous power-law scaling, the species-area relationship, has formed the foundation of many conservation strategies. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because the scale-invariant complexity of fractal ecosystem structure may drive ecological dynamics in space. Here, we use theory and analysis of extensive fish community data from two distinct geographic regions (Hokkaido Island in Japan and the midwestern United States) to show that ecosystem size and complexity play equally-important roles in scaling biodiversity in rivers. The recurrent merging of various streams and rivers forms fractal branching systems, where the prevalence of branching (ecosystem complexity) represents a macro-scale control of the ecosystem's habitat heterogeneity. In the meantime, ecosystem size dictates metacommunity size and total habitat diversity, two factors regulating biodiversity in nature. Our theory predicted that, regardless of simulated species' traits, larger and more branched 'complex' networks supported greater species richness due to increased space and environmental heterogeneity. Both relationships were linear on a log-log scale, indicating a general power-law scaling along the two orthogonal dimensions of ecosystem structure, i.e., size and complexity. In support of this theoretical prediction, the dual power-law scaling of riverine fish diversity has consistently emerged across the study regions despite hosting different fauna with distinct evolutionary histories. The dual scaling of biodiversity may be a pervasive emergent property of branching networks with important implications for biodiversity conservation.
